---
title: "Blast population identification `r params$sample_name`"
author: 'User ID: `r Sys.getenv("USER")`'
date: "`r Sys.Date()`"
output:
  html_document:
    fig_height: 7
    fig_width: 9
    keep_md: yes
    toc: yes
    df_print: paged
params:
  out_path: "/gpfs/data/aifantislab/home/nadorb01/projects/ad_ped_AML_v5/results/" # output directory
  sample_name: "ad_ped_AML_v5_res2" # sample name
  metadata_path: "/gpfs/data/aifantislab/home/nadorb01/projects/ad_ped_AML_v5/results/04_metadata_res.2.csv" # metadata that includes UMAP and clustering information, cell (colnames of seurat_obj) are annotated as "cell" in metadata, aml column contain "AML/Control" information in our case
  metadata_CNV: "/gpfs/data/aifantislab/home/nadorb01/projects/ad_ped_AML_v5/results/infercnv-broad-cell-type-per-patient/CNV_metadata_all_patients.csv" # CNVs to plot for each patient, path to cnv file - columns are sample names, rows CNVs (names matching map_metadata_from_infercnv.txt file columns) that should be plotted for each patient
  CNV_data_path: "/gpfs/data/aifantislab/home/nadorb01/projects/ad_ped_AML_v5/results/infercnv-broad-cell-type-per-patient-res.2/" # inferCNV directory - this looks for 'map_metadata_from_infercnv.txt' files, which can be generated with infercnv::add_to_seurat() function
  umap_val: "UMAPndim30nn25lognorm" # UMAP that your clustering is based - coordinates must be part of metadata
---

```{r setup, include=FALSE}
attach(params) 
library(ggplot2)
library(RColorBrewer)
library(future)
library(dplyr)
library(tibble)
library(stringr)
set.seed(1234)

knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE, 
                      fig.path = file.path(out_path, paste0("05_Malignant-ME-identification-", sample_name, "/")))

# set out_path and create dir 
out_path <- paste0(out_path, "05_Malignant-ME-identification-", sample_name, "/")
if (!dir.exists(out_path)) {dir.create(out_path)}

# additional functions:
shuf <- function(df){
  return(df[sample(1:dim(df)[1], dim(df)[1]),])
}
```
## Load metadata

```{r load_data}
metadata <- read.csv(metadata_path, stringsAsFactors=F)
```

## CNV metadata
We have CNV analysis runs for all samples. Not all samples have annotated CNVs/abnormal karyotype, but some samples show karyotype abnormalities without
annotation. Plot all possible CNVs for each sample and use any cell that carries a CNV in the malignant annotation. 
We know that in samples with annotated CNVs, those cells form patient-specific clusters. If CNVs in samples overlay with sample specific
clusters, they might be worthwhile including.

```{r add_cnv_data}
# combine InferCNV output data
dirs <- dir(CNV_data_path, "map_metadata_from_infercnv.txt", recursive=TRUE, full.names = TRUE)
write.table(dirs, paste0(out_path, "files_used_for_CNV.txt"), quote=FALSE)

write.table(dirs, paste0(out_path, "files_used_for_CNV.txt"), quote=FALSE)

cnv_data <- list()
for (i in seq_along(dirs)){
    cnv_data[[i]] <- read.delim(dirs[[i]], header=TRUE, row.names=1, stringsAsFactors=FALSE)
    cnv_data[[i]] <- cnv_data[[i]][which(rownames(cnv_data[[i]]) %in% metadata$cell),]
}

cnv_data_combined <- cnv_data %>%
                        purrr::reduce(rbind)

print(dim(cnv_data_combined))

# combine CNV data with seurat_obj metadata:
metadata <- merge(metadata, cnv_data_combined, by.x="cell", by.y=0, all.x=T) 

```

# Plot potential CNVs for all patients
Plot all individual CNVs and combine cells that are CNV positive in a common annotation 'CNV_pos'. We are potentially removing AML033, due to small cell number and AML3133
due to lack of identifyable CNVs/patient-specific cluster. 

```{r plot_CNV_data_all_patients}
temp_out <- paste0(out_path, "CNV_plotting/")
dir.create(temp_out)

metadata_cnv <- read.csv(metadata_CNV, stringsAsFactors=F)

# all sample names are stored in my "samples" column, my control samples are called "ControlXX", here I am inititing a new column
# where the Controls are annotated as CNV- before iterating through the patients

metadata$CNV_pos <- ifelse(metadata$samples %in% str_subset(unique(metadata$samples), "Control"), "Control", "CNV-")

for (p in colnames(metadata_cnv)[2:length(colnames(metadata_cnv))]){

    print(p)
    temp_dir <- paste0(temp_out, p, "/")
    dir.create(temp_dir)
    
    if(sum(metadata_cnv[,eval(p)])>0 ){
      p_cnv <- metadata_cnv %>% filter(!!sym(p) ==1) %>% .$CNVs

      for (c in p_cnv){
          # collect information on all CNV+ cells in data set
          metadata$CNV_pos <- ifelse(metadata[,eval(c)] == 1 & metadata$samples == eval(p), "CNV+", metadata$CNV_pos)

          # plot individual CNVs
          metadata$cnv_temp <- ifelse(metadata$samples %in% str_subset(unique(metadata$samples), "Control"), "Control",
                          ifelse(metadata[,eval(c)] == 1 & metadata$samples == eval(p), "CNV+",
                          ifelse(metadata[,eval(c)] == 0 & metadata$samples == eval(p), "CNV-", "AML")))
          metadata$cnv_temp <- factor(metadata$cnv_temp, levels=c("AML","Control","CNV-","CNV+"))
          metadata <- metadata[order(metadata$cnv_temp, decreasing = FALSE),]

          print(ggplot(shuf(metadata), aes_string(paste0(umap_val, "_1"), paste0(umap_val, "_2"), color = "cnv_temp")) +
                      geom_point(size = 0.05) +
                      xlab("UMAP 1") +
                      ylab("UMAP 2") +
                      theme_classic() +
                      ggtitle(paste0(p," ",c))+
                      scale_color_manual(values=c("lightgrey","#606060", "orange", "darkred")) +
                      guides(color=guide_legend(override.aes=list(size=3))) +
                      theme(text = element_text(size=24), panel.grid.major = element_blank(), panel.grid.minor = element_blank()) )

          ggsave(filename = paste0("UMAP_",p,"_",c,"_all_samples.png"), path = temp_dir, width =14 ,height =12 ,device = "png")

          print(ggplot(shuf(metadata %>% filter(samples %in% c(str_subset(unique(metadata$samples), "Control"), eval(p)))), aes_string(paste0(umap_val, "_1"), paste0(umap_val, "_2"), color = "cnv_temp")) +
                      geom_point(size = 0.05) +
                      xlab("UMAP 1") +
                      ylab("UMAP 2") +
                      theme_classic() +
                      ggtitle(paste0(p," ",c))+
                      scale_color_manual(values=c("#606060", "orange","darkred")) +
                      guides(color=guide_legend(override.aes=list(size=3))) +
                      theme(text = element_text(size=24), panel.grid.major = element_blank(), panel.grid.minor = element_blank()) )

          ggsave(filename = paste0("UMAP_",p,"_",c,".png"), path = temp_dir, width =14 ,height =12 ,device = "png")

      }
    }
}


# plot all cells with specific CNVs
for (c in metadata_cnv$CNVs) {
    metadata[,eval(c)] <- as.factor(metadata[,eval(c)])
        print(ggplot(shuf(metadata), aes_string(paste0(umap_val, "_1"), paste0(umap_val, "_2"), color = eval(c))) +
                    geom_point(size = 0.05) +
                    xlab("UMAP 1") +
                    ylab("UMAP 2") +
                    theme_classic() +
                    ggtitle(c)+
                    scale_color_manual(values=c("lightgrey", "darkred")) +
                    guides(color=guide_legend(override.aes=list(size=3))) +
                    theme(text = element_text(size=24), panel.grid.major = element_blank(), panel.grid.minor = element_blank()) )

        ggsave(filename = paste0("UMAP_",c,"_all_samples.png"), path = temp_out, width =14 ,height =12 ,device = "png")

}


# plot all CNV+ cells on UMAP
metadata$CNV_pos <- factor(metadata$CNV_pos, levels=c("Control", "CNV-", "CNV+"))
print(ggplot(shuf(metadata), aes_string(paste0(umap_val, "_1"), paste0(umap_val, "_2"), color = "CNV_pos")) +
            geom_point(size = 0.05) +
            xlab("UMAP 1") +
            ylab("UMAP 2") +
            theme_classic() +
            ggtitle("All CNV+ cells")+
            scale_color_manual(values=c("#606060", "orange","darkred")) +
            guides(color=guide_legend(override.aes=list(size=3))) +
            theme(text = element_text(size=24), panel.grid.major = element_blank(), panel.grid.minor = element_blank()) )

ggsave(filename = paste0("UMAP_all_CNV+_cells_all_samples.png"), path = temp_dir, width =14 ,height =12 ,device = "png")

write.csv(metadata, paste0(out_path, "metadata_cnv.csv"))

```

# Patient occupancy scoring
We want to identify those clusters that are mainly occupied by specific patients. We selected cluster resolution res.0.6 and an occupancy score > 0.7 as the cut off for patient-specific clusters.
Plot those clusters with the patient-specific annotation.

## Define occupancy for each individual patient
Instead of using all AML patients vs Controls together, let's subset to each individual patient+ Controls and calculate occupancy for each cluster.
Call those clusters malignant where we have strong enrichment for the specific patient. 

```{r occupancy_scoring3}

# loop through each cluster in each of the selected resolutions
res <- "res.2"
#res <- c("res.0.4","res.0.5", "res.0.6", "res.0.7", "res.0.8", "res.0.9", "res.1", "res.1.1", "res.1.2", "res.1.3", "res.1.4", "res.1.5" )
for (r in res){
    print(r)
    temp_path <- paste0(out_path, r, "/")
    if(!dir.exists(temp_path)){dir.create(temp_path)}

  # 
  # run loop - subset data to controls+ individual patient 
  samples <- unique(str_replace(metadata$samples, "Control\\d.*", "Control"))
  common_table <- tibble()
  for (s in samples){
  cluster_table <- metadata %>% 
                    mutate(samples_control = str_replace(samples, "Control\\d.*", "Control")) %>%
                    filter(samples_control %in% c(eval(s), "Control")) %>%
                    group_by(!!sym(r)) %>%
                    dplyr::count(aml) %>% 
                    tidyr::spread(aml, n) %>%
                    tidyr::drop_na(AML) %>%
                    tidyr::replace_na(list(Control=0)) %>%
                    dplyr::summarise(sample=eval(s), group_sum=AML+Control, sample_count=AML,  occupancy_score_sample=sample_count/group_sum)

  common_table <- rbind(common_table,cluster_table)
  }

  common_table <- as.data.frame(common_table)
  common_table$cluster_sample <- paste0(as.character(common_table[,eval(r)]), "_",  common_table[,"sample"])
  write.csv(common_table, paste0(temp_path, "occupancy_score_calc_by_patient_",r,".csv"))

    # merge metadata with occupancy score
    metadata$cluster_sample <- paste0(as.character(metadata[,eval(r)]), "_", metadata[,"samples"])
    meta <- full_join(metadata, common_table[3:6], by="cluster_sample")

    plotlist <- list()
    # # plot occupancy score

    # #pre-define cluster colors
    getPalette1 = colorRampPalette(c(brewer.pal(11, "Spectral"), brewer.pal(11, "BrBG")))
    cluster_colors = getPalette1(length(unique(meta[,eval(r)])))
    names(cluster_colors) <- levels(factor(meta[,eval(r)]))

    meta$cluster <- factor(meta[,eval(r)])
    plotlist[[1]] <- ggplot(meta, aes_string(paste0(umap_val,"_1"), paste0(umap_val,"_2"), color = "cluster")) +
                        geom_point(size = 0.01) +
                        xlab("UMAP 1") +
                        ylab("UMAP 2") +
                        theme_classic() + ggtitle(paste0("Clusters ",r))+
                        scale_color_manual(values=cluster_colors) +
                        guides(color=guide_legend(override.aes=list(size=3))) +
                        theme(text = element_text(size=20), panel.grid.major = element_blank(), panel.grid.minor = element_blank())
    print(plotlist[[1]])
    ggsave(filename = glue("UMAP_clusters_{r}.pdf"), plot=plotlist[[1]], path = temp_path, width =13 ,height =10 ,device = "pdf", useDingbats=F)


    # plot occupancy score
    plotlist[[2]] <- ggplot(shuf(meta), aes_string(paste0(umap_val,"_1"), paste0(umap_val,"_2"), color = "occupancy_score_sample")) +
                        geom_point(size = 0.01) +
                        xlab("UMAP 1") +
                        ylab("UMAP 2") +
                        theme_classic() +
                        scale_color_gradient(low = 'grey', high = 'darkblue') + ggtitle("Occupancy score patient-vs-control")+
                        theme(text = element_text(size=20), panel.grid.major = element_blank(), panel.grid.minor = element_blank())
    print(plotlist[[2]])
    ggsave(filename = glue("UMAP_occupancy_score_patient_vs_control_{r}.pdf"), plot=plotlist[[2]], path = temp_path, width =13 ,height =10 ,device = "pdf", useDingbats=F)

    # plot cluster assignment

    meta$Cluster_assign_patient_0.7 <- ifelse(meta$occupancy_score_sample > 0.7, "malignant", "microenvironment")
    meta <- meta %>% tidyr::replace_na(list(Cluster_assign_patient_0.7 ="Control"))
    plotlist[[3]] <- ggplot(shuf(meta), aes_string(paste0(umap_val,"_1"), paste0(umap_val,"_2"), color = "Cluster_assign_patient_0.7")) +
                        geom_point(size = 0.01) +
                        xlab("UMAP 1") +
                        ylab("UMAP 2") +
                        theme_classic() + ggtitle(paste0("Clusters ",r, " 0.7 cut off"))+
                        scale_color_manual(values=c("grey","darkred","black")) +
                        guides(color=guide_legend(override.aes=list(size=3))) +
                        theme(text = element_text(size=12), panel.grid.major = element_blank(), panel.grid.minor = element_blank())
    
    print( plotlist[[3]])
    ggsave(filename = glue("UMAP_cluster_assignment_patient_vs_control_0.7_{r}.pdf"), plot=plotlist[[3]], path = temp_path, width =7 ,height =5 ,device = "pdf", useDingbats=F)



    print(cowplot::plot_grid(plotlist=plotlist,ncol=3, nrow=1))
    ggsave(filename = glue("UMAPs_{r}_occupancy_combined.pdf"), path = temp_path, width =23 ,height =7 ,device = "pdf", useDingbats=F)


    write.csv(meta, paste0(temp_path, "metadata_",r,".csv"))
}

# uses the last run resolution downstream (re-running the entire script with the correct resolution is required)
metadata <- meta 

```


# Is a given cell malignant or part of the microenvironment?
We are first selecting all cells that do belong to a patient-specifc cluster as defined by the occupancy score, and then use the CNV annotations
to add an additional layer of information to also select cells that do not occupy patient-specific space. We are basically combining "res.0.5_cluster_assign_0.7" and
"CNV_pos" in this step.

```{r malignant_assignment_patient_vs_control}

metadata$samples <- factor(metadata$samples)


metadata$which_malignant <- ifelse( metadata$Cluster_assign_patient_0.7=="malignant" & metadata$CNV_pos=="CNV+", "CNV_occupancy_score",
                            ifelse( metadata$Cluster_assign_patient_0.7=="malignant", "occupancy_score",
                            ifelse(metadata$CNV_pos =="CNV+", "CNV_analysis", "none" )))

metadata$malignant <- ifelse(metadata$samples %in% str_subset(unique(metadata$samples), "Control"), "Control",
                      ifelse( metadata$Cluster_assign_patient_0.7=="malignant" | metadata$CNV_pos=="CNV+", "malignant", "microenvironment"))

metadata$malignant <- factor(metadata$malignant, levels=c("Control", "malignant", "microenvironment"))


ggplot(shuf(metadata), aes_string(paste0(umap_val, "_1"), paste0(umap_val, "_2"), color = "malignant")) +
            geom_point(size = 0.05) +
            xlab("UMAP 1") +
            ylab("UMAP 2") +
            theme_classic() +
            ggtitle("Malignant_microenvironment")+
            scale_color_manual(values=c("#606060","darkred", "orange")) +
            guides(color=guide_legend(override.aes=list(size=3))) +
            theme(text = element_text(size=24), panel.grid.major = element_blank(), panel.grid.minor = element_blank()) 
ggsave(filename = paste0("UMAP_control_malignant_ME_patient_wise_os.pdf"), path = out_path, width =14 ,height =12 ,device = "pdf", useDingbats=F)

```


```{r save}
write.csv(metadata, paste0(out_path, "metadata_malignant_ME.csv"))
```

